The markings of the trade: exploring the patterns of olive oil production in Roman Baetica

### 2020-02-12 Paper/jaccardbaetica.tex
new changes bib

@@

The aim of this study is to detect the patterns of olive oil production that link amphora workshops and amphoric stamps. Our case of study has been focused on the production processes located in Baetica province (currently Andalusia) from 1st to 3rd AD. In particular, we want to explore economic dynamics that include the production and distribution of olive oil trade. 
Baetica became important production and distribution centres during the Roman Empire. However, it remains under debate how this province was organized and whether it is possible to identify patterns in the olive oil market. Amphoric stamps are used to identify the presence of different producer groups that might share similar stamps. To achieve this goal, we analyse a set of stamps from different workshops in Baetica province in order to detect a relation between the distribution of amphoric stamps and the economic structure in this province. Here we use methods borrowed from Ecology that allow us to identify if amphora workshops share similar amphoric stamps depending on the spatial distance.

The analysis explores how quantitative approach provides a useful tool for the interpretation of the economic processes. Finally, results pretend to highlight the organization of Baetican olive oil production in the Roman Empire linked to the differences observed in the archaeological evidence.

\xavi{el abstract es demasiado espécifico al inicio y demasiado genérico después en relación a lo que se ha hecho}

@@

Despite this issue has been widely discussed over many decades, the understanding about the production processes is still under debate, mostly due to the lack of written records. The application of different approaches combined with the archaeological evidence has allowed us partly to deal with the complexity of understanding the Roman production \citep{orengo_seeds_2016,brughmans_roman_2016,coto-sarmiento_maria_bayesian_????}.

@@

\xavi{guai el mapa pero el mapa pequeño debería ser como mínimo a nivel de Península Ibérica y, a ser posible, el imperio con provincias?}

The chronology in the workshops is widely diverse from the first to the third centuries AD \citep{millet_anforas_1998,rodriguez_baetican_1998,chic2005comercio}. Some stamps show a more specific chronology while the majority of them display a large activity of production that it can be difficult for specifying an accurate chronology. This could be due to two reasons; firstly, most of the workshops were partially excavated and focused on archaeological surveys in order to collect the maximum stamps as possible; secondly, Dressel 20 was produced during almost three centuries with apparently few changes. (podría citar nuestro paper por toda la face).

@@

From the database, we collected the site where stamps were found and the stamp code. We also created a new column with the area of the stamps. These areas, known as \textit{conventus}, were administrative centres for territorial organization in the Roman Empire. Dressel 20 stamps were found in three different \textit{conventus}: \textit{Hispalensis} (currently Seville, hereafter Hispalis), \textit{Cordubensis} (currently C\'ordoba, hereafter Corduba) and \textit{Astigi} (currently Écija, Sevilla, hereafter Astigi) \citep{rodriguez_economioleicola_1977,chic_datos_2001,berni_millet_epigrafianforica_2008} .

@@

The frequency of stamps per workshop can be seen in Fig. \ref{stamps}. Most stamps are distributed in only one workshop whereas few workshops concentrate a high frequency of different amphoric stamps. The type of distribution is also frequent in Roman economy where we observed a self-organized complex system patterns typical of a free market \citep{coto-sarmiento_maria_bayesian_????}.

@@

It is well known that this product was not frequently consumed by autonomous population. The absence of olive oil importation before the conquest can be reflected by the shortage of this product until the arrival of high percentage of Dressel 20 amphorae to Britannia \citep[ 1]{carreras_britannia_1998}. Additionally, the land of Britannia showed inadequate for the olive oil production due to low environmental conditions. This issue was necessary for creating an important export apparatus from Baetica to supply Roman legions.
In this moment, we detect an increase of the olive oil exportation concurring with the displacement of Roman legions during the military campaigns \citep[ 161]{monfort_britanniaen_1998}. This fact will be spatial incidence in sites close at Hadrian Wall.  
Olive oil production in Baetica would cross the Atlantic until they reached the province and redistribute throughout the area from a series of strategic points \citep{carreras_atlantic_2012}. The increase of the exportation of Dressel 20 amphorae created an important commercial network for exchanges. Thus, the network was mainly focused on the support of soldiers during military campaigns. 
%debate intercambios puntuales o redes comerciales de abastecimiento

@@

The similarity of stamps show a results less significative than Britannia province as it can be seen in the dendrogram Fig. \ref{germap}.

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/dendroger5.pdf}
\caption{Dendrogram obtained by Morisita-Horn algorthm of different sites in Germania. Colours are represented by type of sites divided into military sites (red), civil sites (green) and no specific (grey)}
\label{germap}
\end{figure}

@@

The results of the case study could be interpreted by several reasons. On the one hand, the use of these amphoric stamps could have been exclusively running by the owner or owners of the workshop to distinguish the amphora workshop (CITAR). This hypothesis would explain the fact that we do not find similar stamps in different workshops, however, we found different amphora stamps in the same workshops that they would be barely difficult to assign different owners. Neither a quality distinction does seem to be used to specify the value of the product \citep{callender}. On the other hand, it could be interpreted somehow a batch systematic organization. Considering that Dressel 20 was not marked in most cases, some authors point out that potters marked amphorae to prepare and distribute the commodity to be shipped \citep{berni_millet_epigrafianforica_2008}. This method would be used as an identifier to count the number of amphorae of a branch (CITAR). This method could also have served to identify different groups of potter workers working in the same amphora workshop. Potters could have marked the amphorae to distinguish different groups working in parallel \citep{li_crossbows_2014}. This it would explain wherefore why we detect different stamps in a same workshop. 
In any case, we do not have enough archaeological evidence that can validate the interpretations presented here and our results are certainly valid only with the context of our case study.

>>>>

\author[ceipacadress]{Maria Coto-Sarmiento\corref{mycorrespondingauthor}}
\cortext[mycorrespondingauthor]{Corresponding author}
\ead{mcotsar@gmail.com}

\author[edadress]{Xavier Rubio-Campillo}

\address[edadress]{School of History, Classic \& Archaeology, Room OOM.33, William Robertson Wing, Old Medical School, Teviot Place, University of Edinburgh, UK}
\address[ceipacadress]{CEIPAC, Department of Prehistory and Archaeology, Montalegre, 6-8, 08001, University of Barcelona, Barcelona, Spain}

\begin{keyword}
Roman Empire; amphora workshops; Dressel 20; social learning; cultural evolution
\end{keyword}

@@

The aim of this study is to explore economics dynamics in the production and distribution of olive oil trade. 
Our case of study has been focused on the production processes located in Baetica province (currently Andalusia) from 1st to 3rd AD. In particular, we want to detect patterns of olive oil production that link amphora workshops and amphoric stamps. Baetica became important production and distribution centres during the Roman Empire. However, it remains under debate how this province was organised and whether it is possible to identify patterns in the olive oil market. Amphoric stamps are used to identify the presence of different producer groups that might share similar stamps. To achieve this goal, we analyse a set of stamps from different workshops in Baetica province in order to detect a relation between the distribution of amphoric stamps and the economic structure in this province. Here we use methods borrowed from Ecology that allow us to identify if amphora workshops share similar amphoric stamps depending on the spatial distance.

The analysis explores how quantitative approach provides a useful tool for the interpretation of the economic processes. Finally, results pretend to highlight the organisation of Baetican olive oil production in the Roman Empire linked to the differences observed in the archaeological evidence.

@@

Despite this issue has been widely discussed over many decades, the understanding about the production processes is still under debate, mostly due to the lack of written records. The application of different quantitative approaches combined with the archaeological evidence has allowed us partly to deal with the complexity of understanding the Roman production \citep{orengo_seeds_2016,
brughmans_roman_2016,coto-sarmiento_identifying_2018,bayesian_2018}.

@@

The chronology in the workshops is widely diverse from the first to the third centuries AD \citep{millet_anforas_1998,rodriguez_baetican_1998,chic2005comercio}. Some stamps show a more specific chronology while the majority of them display a large activity of production that it can be difficult for specifying an accurate chronology. This could be due to two reasons; firstly, most of the workshops were partially excavated and focused on archaeological surveys in order to collect the maximum stamps as possible; secondly, Dressel 20 was produced during almost three centuries with apparently few changes \citep{berni_piero_chapter_2017}.

@@

From the database, we collected the site where stamps were found and the stamp code. We also created a new column with the area of the stamps. These areas, known as \textit{conventus}, were administrative centres for territorial organization in the Roman Empire. Dressel 20 stamps were found in three different \textit{conventus}: \textit{Hispalensis} (currently Seville, hereafter Hispalis), \textit{Cordubensis} (currently C\'ordoba, hereafter Corduba) and \textit{Astigi} (currently Écija, Sevilla, hereafter Astigi) \citep{rodriguez_economioleicola_1977,
chicdatos2001,berni_millet_epigrafianforica_2008} .

@@

The frequency of stamps per workshop can be seen in Fig. \ref{stamps}. Most stamps are distributed in only one workshop whereas few workshops concentrate a high frequency of different amphoric stamps. The type of distribution is also frequent in Roman economy where we observed a self-organized complex system patterns typical of a free market \citep{bayesian_2018}.

@@

It is well known that this product was not frequently consumed by autonomous population. The absence of olive oil importation before the conquest can be reflected by the shortage of this product until the arrival of high percentage of Dressel 20 amphorae to Britannia \citep[ 1]{carreras_britannia_1998}. Additionally, the land of Britannia showed inadequate for the olive oil production due to low environmental conditions. This issue was necessary for creating an important export apparatus from Baetica to supply Roman legions.
In this moment, we detect an increase of the olive oil exportation concurring with the displacement of Roman legions during the military campaigns \citep[161]{monfort_britanniaen_1998}. This fact will be spatial incidence in sites close at Hadrian Wall.  
Olive oil production in Baetica would cross the Atlantic until they reached the province and redistribute throughout the area from a series of strategic points \citep{carreras_atlantic_2012}. The increase of the exportation of Dressel 20 amphorae created an important commercial network for exchanges. Thus, the network was mainly focused on the support of soldiers during military campaigns. 
%debate intercambios puntuales o redes comerciales de abastecimiento

@@

The similarity of stamps show a results less significant than Britannia province as it can be seen in the dendrogram (Fig. \ref{germap}).

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/dendroger5.pdf}
\caption{Dendrogram obtained by Morisita-Horn algorthm of different sites in Germania. Colours are represented by types of sites divided into military sites (red), civil sites (green) and no specific (grey)}
\label{germap}
\end{figure}

@@

The results of the case study could be interpreted by several reasons. On the one hand, the use of these amphoric stamps could have been exclusively running by the owner or owners of the workshop to distinguish the amphora workshop (CITAR). This hypothesis would explain the fact that we do not find similar stamps in different workshops; however, we found different amphora stamps in the same workshops that they would be barely difficult to assign different owners. Neither a quality distinction does seem to be used to specify the value of the product \citep{callenderindex}. On the other hand, it could be interpreted somehow a batch systematic organisation. Considering that Dressel 20 was not marked in most cases, some authors point out that potters marked amphorae to prepare and distribute the commodity to be shipped \citep{berni_millet_epigrafianforica_2008}. This method would be used as an identifier to count the number of amphorae of a branch \citep{juanmorostesis}. This method could also have served to identify different groups of potter workers working in the same amphora workshop. Potters could have marked the amphorae to distinguish different groups working in parallel \citep{li_crossbows_2014}. This it would explain wherefore why we detect different stamps in a same workshop. 
In any case, we do not have enough archaeological evidence that can validate the interpretations presented here and our results are certainly valid only with the context of our case study.
